Structured Genealogy Processes

### 2022-02-18 ms.Rnw
thm2 update

\paragraph{Deaths and population size}
The population, the set of infected individuals in the context of epidemiology, is structured, categorized into interchangeable demes. 
We can then define population size function and death functions in Deme $j$, respectively, as $I^{j}, D_{j}:\Xspace\to\N$ for any $j=1,\cdots,J$. 
\begin{equation}
  \jumprate_u(t,x) > 0 \Longrightarrow I^j(x+u)-I^j(x)=\sum_{j'}\,B_{j',j}(u)+\sum_{j'}M_{j',j}(u)-D_j(u)-\sum_{j'}M_{j,j'}(u),
\end{equation}

for all $x, u\in\Xspace$. We disallow simultaneous death at a single death event at this stage and define the set of death events in Deme $j$ as $\Dset^j\coloneqq\,D_j^{-1}(\{1\})$. 
We also insist, birth and death events don't coincide for any demes, $j,j'$, that is,  $\Bset_{j}\cap\Dset_{j'}=\emptyset$. 
Eventually, we can have the function for total population size by letting $I(x)\coloneqq\sum_{j=1}^J\,I^j(x)$ and it's trivial that

@@

\begin{equation}
  \mathrm{inven}(\omega,j)\coloneqq
  \begin{cases}
    \mathrm{initialize}(I^j(\sg{X}_0(\omega)), & \text{if }K(\omega)=0; \\
    \mathrm{add}(\mathrm{inven}(\predec{\omega},j)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\Bset_j;\\
    \mathrm{drop}(\mathrm{inven}(\predec{\omega},j),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\Dset_j; \\
    \mathrm{movein}(\mathrm{inven}(\predec{\omega},j),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\bigcup_i\Mset_{ij}; \\
    \mathrm{drop}(\mathrm{inven}(\predec{\omega},j),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\bigcup_i\Mset_{ji}; \\
    \mathrm{inven}(\predec{\omega},j), & \text{otherwise}.
  \end{cases}
\end{equation}

@@

We can further define the deme proportion function $\demefuc(t,\omega)\coloneqq\{\demefuc^j(t,\omega)\}_{j=1}^J$, where $\demefuc^j(t,\omega)\coloneqq\frac{\vert\Inv_t^j(\omega)\vert}{\vert\Inv_t(\omega)\vert}$.

@@

\begin{equation*}
  \begin{aligned}
    % P_{\Och_q|\Och_{q-1},\Hist}(o_q|o_{q-1},h)&=\prod_{k=1}^K\sum_{n_k\in\Omega_{qk}}\auxmeas_{u_k,x_k-u_k}(n_k)=\prod_{k=1}^K\sum_{n_k\in\Omega_{qk}}\left(\sum_{j=1}^J\dfrac{\indicator{\Inv^j_{t_{k-1}}(\Omega_q)}(n_k)}{I_j(x_k-u_k)}\right).
    P_{\Och_q|\Och_{q-1},\Hist}(o_q|o_{q-1},h) &= \prod_{k=1}^K\sum_{\omega\in\Omega_q}\auxmeas_{u_k,x_k-u_k}(N_k(\omega)) = \prod_{k=1}^K\sum_{\omega\in\Omega_q}\left(\sum_{j=1}^J\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I^j(x_k-u_k)}\right),
  \end{aligned}
\end{equation*}

@@

We define $\phi_{qk}^{j}\coloneqq\sum_{\omega\in\Omega_q}\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I^j(x_k-u_k)}$, 
and let $I_k^j=I^j(x_k)$, $\ell_{qk}^{j}=\ell^j(t_k, \Och_{q-1})$.
Note that $\ell^j_{1k}=0$ for all $k$ and $j$.

@@

  % B, j!=j'
  \item There was a birth event between two different demes, that is $u_k\in\Bset^{j,j'}$ and $\delta_{j,j'}=0$:
    \begin{enumerate}
      % There were  l'*(I-l) possibilities of transitions
      \item and given that it was presented as a deme-transition event in the pre-existing genealogy, that is, $t_k\in\Zset^{j,j'}(\Och_{q-1})$:
        \begin{enumerate}
          % the genealogy has the exact l' lineage as the child, and the sample is the exact 1 as the parent
          % 1/[l'*(I-l)] * l' * 1/1
          \item if the $q$-th sample was from deme $j$ and is exactly the parent in the birth, that is, $t_k\in\Cset^{j,j'}(\Och_q)$ and $\delta_{j,j''}=1$, then $\phi_{qk}^{j''}=\delta_{j,j''}/\left(I_k^{j}-\ell_{qk}^{j}\right)$;
          \item otherwise, $t_k\in\Zset^{j,j'}(\Och_q)$,and $\phi_{qk}^{j''}=1-\delta_{j,j''}/\left(I_k^{j}-\ell_{qk}^{j}\right)$;
        \end{enumerate}
      % there were in total I*I' in the birth, while l*l' is excluded since it was not presented as coalescent, and (I-l)*l' are excluded which as presented as transition. == I I' - l l' - (I-l) l' = I (I'-l')
      \item or given that it was not presented in $\Och_{q-1}$, \ie $t_k\notin\Zset^{j,j'}(\Och_{q-1})\cup\Cset^{j,j'}(\Och_{q-1})$:
        \begin{enumerate}
          % the genealogy has l lineages as parent, and we choose one of them 1/l: 1/[I (I'-l')] * l * 1/l
          \item if the $q$-th sample was from deme $j'$ and is coalesced at $t_k$, \ie $t_k\in\Cset^{j,j'}(\Och_q)$ and $\delta_{j',j''}$, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j}(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
          % 
          \item if the $q$-th sample was from deme $j'$ and is presented as a deme-transition event in $\Och_q$, \ie $t_k\in\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j}(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;

          \item otherwise, $t_k\notin\Cset^{j,j'}(\Och_q)\cup\Zset^{j,j'}(\Och_q)$, and $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left(I_{k}^{j'}-\ell_{qk}^{j'}\right)$;
        \end{enumerate}
    \end{enumerate}
\end{enumerate}

@@

    % M
    1, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ t_k\in\Zset^{j,j'}(\Och_{q}),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
    \frac{\delta_{j',j''}}{(I_{k}^j-\ell_{q,k}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
    1-\frac{\delta_{j',j''}}{(I_{k}^j-\ell_{q,k}^{j})(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]

@@

  Consider $u_k\in\Mset^{j,j'}$. Let $q^*=\{q\;\vert\;t_k=\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1})\}$ and $\ell^*=\ell_{q^*k}^{j''_{q^*k}}$. We then define $Q_k=\{q\;\vert\;j''_{qk}=j'\ \text{and}\ q<q^*\}$ as an ordered set and denote $[Q_k]_m$ as the $m$-th element in $Q_k$, where, $m=1,\dots, \ell^*$. 
  \begin{enumerate}[(a)]
    \item If $q<q^*$, then $t_k\notin\Zset^{j,j'}(\Och_q)$. We will have
      \begin{equation*}
        \Phi_k = \prod_{q\in Q_k} \left(1-\frac{1}{(I_k^j-\ell_{qk}^j)(I_k^{j'}-\ell_{qk}^{j'})}\right).
      \end{equation*}
      Note that, $\{\ell_{qk}^{j'}\;\vert\;q\in Q_k\}=\{0,1,\dots,\ell^*-1\}$ then the previous equation becomes
      \begin{equation*}
        \Phi_k = \prod_{m=1}^{\ell*} \left(1-\frac{1}{\mathcal{Y}_m (I_k^{j'}-m+1)}\right) = \prod_{m=1}^{\ell*} \frac{I_k^{j'}-m+1-(\mathcal{Y}_m)^{-1}}{I_k^{j'}-m+1},
      \end{equation*}
      where $\mathcal{Y}_m\coloneqq I_k^{j}-\ell_{[Q_k]_m k}^{j}$.
  \end{enumerate}

>>>>

\paragraph{Deaths and population size}
The population, the set of infected individuals in the context of epidemiology, is structured, categorized into interchangeable demes. 
We can then define population size function and death functions in Deme $j$, respectively, as $I^{j}, D_{j}:\Xspace\to\N$ for any $j=1,\cdots,J$. 
\begin{equation}
  \jumprate_u(t,x) > 0 \Longrightarrow I_j(x+u)-I_j(x)=\sum_{j'}\,B_{j',j}(u)+\sum_{j'}M_{j',j}(u)-D_j(u)-\sum_{j'}M_{j,j'}(u),
\end{equation}

for all $x, u\in\Xspace$. We disallow simultaneous death at a single death event at this stage and define the set of death events in Deme $j$ as $\Dset^j\coloneqq\,D_j^{-1}(\{1\})$. 
We also insist, birth and death events don't coincide for any demes, $j,j'$, that is,  $\Bset_{j}\cap\Dset_{j'}=\emptyset$. 
Eventually, we can have the function for total population size by letting $I(x)\coloneqq\sum_{j=1}^J\,I_j(x)$ and it's trivial that

@@

\begin{equation}
  \mathrm{inven}(\omega,j)\coloneqq
  \begin{cases}
    \mathrm{initialize}(I_j(\sg{X}_0(\omega)), & \text{if }K(\omega)=0; \\
    \mathrm{add}(\mathrm{inven}(\predec{\omega},j)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\Bset_j;\\
    \mathrm{drop}(\mathrm{inven}(\predec{\omega},j),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\Dset_j; \\
    \mathrm{movein}(\mathrm{inven}(\predec{\omega},j),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\bigcup_i\Mset_{ij}; \\
    \mathrm{drop}(\mathrm{inven}(\predec{\omega},j),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\bigcup_i\Mset_{ji}; \\
    \mathrm{inven}(\predec{\omega},j), & \text{otherwise}.
  \end{cases}
\end{equation}

@@

We can further define the deme proportion function $\demefunc(t,\omega)\coloneqq\{\demefunc^j(t,\omega)\}_{j=1}^J$, where $\demefunc^j(t,\omega)\coloneqq\frac{\vert\Inv_t^j(\omega)\vert}{\vert\Inv_t(\omega)\vert}$.

@@

\begin{equation*}
  \begin{aligned}
    % P_{\Och_q|\Och_{q-1},\Hist}(o_q|o_{q-1},h)&=\prod_{k=1}^K\sum_{n_k\in\Omega_{qk}}\auxmeas_{u_k,x_k-u_k}(n_k)=\prod_{k=1}^K\sum_{n_k\in\Omega_{qk}}\left(\sum_{j=1}^J\dfrac{\indicator{\Inv^j_{t_{k-1}}(\Omega_q)}(n_k)}{I_j(x_k-u_k)}\right).
    P_{\Och_q|\Och_{q-1},\Hist}(o_q|o_{q-1},h) &= \prod_{k=1}^K\sum_{\omega\in\Omega_q}\auxmeas_{u_k,x_k-u_k}(N_k(\omega)) = \prod_{k=1}^K\sum_{\omega\in\Omega_q}\left(\sum_{j=1}^J\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I_j(x_k-u_k)}\right),
  \end{aligned}
\end{equation*}

@@

We define $\phi_{qk}^{j}\coloneqq\sum_{\omega\in\Omega_q}\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I_j(x_k-u_k)}$, 
and let $I_k^j=I_j(x_k)$, $\ell_{qk}^{j}=\ell^j(t_k, \Och_{q-1})$.
Note that $\ell^j_{1k}=0$ for all $k$ and $j$.

@@

  % B, j!=j'
  \item There was a birth event between two different demes, 
  that is $u_k\in\Bset^{j,j'}$ and $\delta_{j,j'}=0$:
    \begin{enumerate}
      % There were  l'*(I-l) possibilities of transitions
      \item and given that it was presented as a deme-transition event in the pre-existing genealogy, 
      that is, $t_k\in\Zset^{j,j'}(\Och_{q-1})$:
        \begin{enumerate}
          % the genealogy has the exact l' lineage as the child, and the sample is the exact 1 as the parent
          % 1/[l'*(I-l)] * l' * 1/1
          \item if the $q$-th sample was from deme $j$ and is exactly the parent in the birth, 
          that is, $t_k\in\Cset^{j,j'}(\Och_q)$ and $\delta_{j,j''}=1$, then $\phi_{qk}^{j''}=\delta_{j,j''}/\left(I_k^{j}-\ell_{qk}^{j}\right)$;
          \item otherwise, 
          $t_k\in\Zset^{j,j'}(\Och_q)$,and $\phi_{qk}^{j''}=1-\delta_{j,j''}/\left(I_k^{j}-\ell_{qk}^{j}\right)$;
        \end{enumerate}
      % there were in total I*I' in the birth, while l*l' is excluded since it was not presented as coalescent, and (I-l)*l' are excluded which as presented as transition. == I I' - l l' - (I-l) l' = I (I'-l')
      \item or given that it was not presented in $\Och_{q-1}$, 
      \ie $t_k\notin\Zset^{j,j'}(\Och_{q-1})\cup\Cset^{j,j'}(\Och_{q-1})$:
        \begin{enumerate}
          % the genealogy has l lineages as parent, and we choose one of them 1/l: 1/[I (I'-l')] * l * 1/l
          \item if the $q$-th sample was from deme $j'$ and is coalesced at $t_k$, 
          \ie $t_k\in\Cset^{j,j'}(\Och_q)$ and $\delta_{j',j''}$, 
          then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j}(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
          % I - l can be the parent, and we choose one of them 1/(I - l) as the sample: 1/[I (I'-l')] * (I-l) * 1/(I-l)
          \item if the $q$-th sample was from deme $j'$ and is presented as a deme-transition event in $\Och_q$, 
          \ie $t_k\in\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j}(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
          \item otherwise, $t_k\notin\Cset^{j,j'}(\Och_q)\cup\Zset^{j,j'}(\Och_q)$, 
          and $$\phi_{qk}^{j''}=1-\delta_{j',j''}/(I_{k}^{j'}-\ell_{qk}^{j'});$$
        \end{enumerate}
    \end{enumerate}
\end{enumerate}

@@

    % M
    1, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ t_k\in\Zset^{j,j'}(\Och_{q}),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
    \frac{\delta_{j',j''}}{(I_{k}^j-\ell_{q,k}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1}),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
    1-\frac{\delta_{j',j''}}{(I_{k}^j-\ell_{q,k}^{j})(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]

@@

  Consider $u_k\in\Mset^{j,j'}$. Let $q^*=\{q\;\vert\;t_k=\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1})\}$ and $\ell^*=\ell_{q^*k}^{j'}$. We then define $Q_k^{j'}=\{q\;\vert\;j''_{qk}=j'\ \text{and}\ q\leq q^*\}$ as an ordered set and denote $[Q_k^{j'}]_m$ as the $m$-th element in $Q_k^j$, where, $m=1,\dots, \ell^*$.
  \begin{enumerate}[(a)]
    \item \label{scn: notZ} If $q<q^*$, then $t_k\notin\Zset^{j,j'}(\Och_q)$. We will have
      \begin{equation*}
        \Phi_k = \prod_{q\in Q_k^{j'}} \left(1-\frac{1}{(I_k^j-\ell_{qk}^j)(I_k^{j'}-\ell_{qk}^{j'})}\right).
      \end{equation*}
      Note that, $\{\ell_{qk}^{j'}\;\vert\;q\in Q_k^{j'}\}=\{0,1,\dots,\ell^*\}$ then the previous equation becomes
      \begin{equation*}
        \Phi_k = \prod_{m=1}^{\ell^*} \left(1-\frac{1}{\mathcal{Y}_m^j (I_k^{j'}-m+1)}\right) = \prod_{m=1}^{\ell^*} \frac{I_k^{j'}-m+1-(\mathcal{Y}_m^j)^{-1}}{I_k^{j'}-m+1},
      \end{equation*}
    where $\mathcal{Y}_m^j\coloneqq I_k^{j}-\ell_{[Q_k^{j'}]_m k}^{j}\in\Z$.

    \item If $q\geq q^*$, then $t_k\in\Zset^{j,j'}$, and obviously, with $\mathcal{Y}_m^j$ defined as before,
      \begin{equation*}
        \Phi_k = \frac{1}{\mathcal{Y}_{\ell^*}^j (I_k^{j'}-\ell^*)}\, \prod_{m=1}^{\ell^*} \frac{I_k^{j'}-m+1-(\mathcal{Y}_m^j)^{-1}}{I_k^{j'}-m+1}.
      \end{equation*}
  \end{enumerate}

  Consider $u_k\in\Bset^{j,j'}$ and $j\neq j'$. Let $q^*=\{q\;\vert\;t_k=\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1})\}$ and $q'=\{q\;\vert\;t_k=\Cset^{j,j'}(\Och_q)\setminus\Cset^{j,j'}(\Och_{q-1})\}$.
  % first, get q* as the sample of deme-transition
  % q' as the sample of coalescent
  \begin{enumerate}[(a)]
    % if not exists q* and q'
    \item If $q^*=\emptyset$ and $q'=\emptyset$, then $t_k\notin\Zset^{j,j'}(\Och_i)\cup\Cset^{j,j'}(\Och_{i})$. So for any $q\leq i$, we have
      \begin{equation*}
        \Phi_k=\prod_{m=0}^{\ell_{ik}^{j'}}\left(1-\frac{1}{I_k^{j'}-m}\right)=1-\frac{\ell_{ik}^{j'}+1}{I_k^{j'}}.
      \end{equation*}
    % if not exists q* but q'
    \item If $q^*\neq\emptyset$ but $q'=\emptyset$, then $t_k\in\Zset^{j,j'}(\Och_{i})$. Let $\ell^*=\ell_{q^*k}^{j'}$ and for $q^*<q\leq i$, we have
      \begin{equation*}
      \begin{split}
        \Phi_k &= \prod_{m=0}^{\ell^*}\left(1-\frac{1}{I_k^{j'}-m}\right)\,\frac{1}{I_k^{j}(I_k^{j'}-\ell^*)}\,\prod_{m=\ell_{q^*k}^{j}}^{\ell_{ik}^{j}}\left(1-\frac{1}{I_k^{j}-m}\right) \\
        &= \frac{I_k^{j'}-\ell^*-1}{I_k^{j'}}\frac{1}{I_k^j(I_k^{j'}-\ell^*)}\frac{I_k^j-\ell_{ik}^{j}-1}{I_k^j-\ell_{q^*k}^{j}}\\
        &= \frac{1}{\mathcal{Y}^{j}_{\ell^*}(I_k^{j'}-\ell^*)}\left(1-\frac{\ell_{ik}^{j}+1}{I_k^j}\right)\left(1-\frac{\ell^*+1}{I_k^{j'}}\right),
      \end{split}
      \end{equation*}
      where $\mathcal{Y}^{j}_{\ell^*}$ is as defined in \ref{scn: notZ} when $u_k\in\Mset^{j,j'}$.
    % if not exists q' but q*
    \item If $q'\neq\emptyset$ but $q^*=\emptyset$, then $t_k\notin\Zset^{j,j'}(\Och_{q'-1})$ and $t_k\in\Cset^{j,j'}(\Och_i)$. Let $\ell'=\ell_{q'k}^{j'}$ and for $q'<q\leq i$, we have
      \begin{equation*}
        \begin{split}
          \Phi_k &= \prod_{m=0}^{\ell'}\left(1-\frac{1}{I_k^{j'}-m}\right)\,\frac{1}{I_k^{j}(I_k^{j'}-\ell')}=\left(1-\frac{\ell'+1}{I_{k}^{j'}}\right)\frac{1}{I_k^{j}(I_k^{j'}-\ell')}.
        \end{split}
      \end{equation*}
    % if exists q* and q', and q' > q*
    \item If $q^*\neq\emptyset$, $q'\neq\emptyset$, and obviously, $q^*<q'$, then $t_k\in\Zset^{j,j'}(\Och_{q^*})$, $t_k\notin\Zset^{j,j'}(\Och_{q'})$, and $t_k\in\Cset^{j,j'}(\Och_{q'})$. Let $\ell^*=\ell_{q^*k}^{j'}$ and $\hat{\ell}=\ell_{q'k}^{j}$. We have
      \begin{equation*}
      \begin{split}
        \Phi_k &= \prod_{m=0}^{\ell^*}\left(1-\frac{1}{I_k^{j'}-m}\right)\,\frac{1}{I_k^{j}(I_k^{j'}-\ell^*)}\prod_{m=\ell_{q^*k}^{j}}^{\hat{\ell}}\left(1-\frac{1}{I_k^{j}-m}\right)\,\frac{1}{I_k^{j}-\hat{\ell}} \\
        &= \frac{1}{\mathcal{Y}_{\ell^*}^{j}(I_k^{j}-\hat{\ell})(I_k^{j'}-\ell^*)}\left(1-\frac{\hat{\ell}+1}{I_k^{j}}\right)\left(1-\frac{\ell^*+1}{I_k^{j'}}\right)
      \end{split}
      \end{equation*}
  \end{enumerate}
